---
title: "Cibersortx Cell Fraction Analysis"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
pkgdown:
  as_is: true
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(glue)
library(readxl)
library(RColorBrewer)
library(cowplot)
library(SummarizedExperiment)
library(DirichletReg)
```

```{r}
set.seed(2026)
```


# Import data

```{r}
load("../data/raw/rnaseq_hersh.Rdata")

subgroup_palette <- c(
  "EZH2"      = "#E69F00", 
  "TET2-bi"   = "#56B4E9", 
  "-7"        = "#009E73", 
  "STAG2"     = "#CC79A7", 
  "del5q-IB"  = "#F0E442", 
  "SF3B1-IB"  = "#0072B2", 
  "Complex"   = "#D55E00", 
  "MDS-LB"    = "#999999", 
  "MDS-IB1"   = "grey40", 
  "MDS-IB2"   = "black"
)

subgroups <- names(subgroup_palette)

who_palette <- c(
  "MDS-5q"    = "#F0E442", 
  "MDS-SF3B1" = "#0072B2", 
  "MDS-TP53"  = "#D55E00", 
  "MDS-LB"    = "#999999", 
  "MDS-IB1"   = "grey40", 
  "MDS-IB2"   = "black"
)

who_levels <- names(who_palette)

sample_metadata <- colData(se_hersh) |> 
  as.data.frame() |> 
  select(-SEX, -AGE, -Exam_Array) |> 
  rename_with(str_to_snake) |> 
  rownames_to_column("sample") |> 
  select(sample, who, bm_blast, wbc, hb, plt, ipssm, sub_group) |> 
  mutate(who = factor(who, levels = who_levels)) |> 
  mutate(sub_group = factor(sub_group, levels = subgroups))

types_palette <- c(
                      "hspc" = "#984EA3", 
          "dc_monocytic_gmp" = "#4DAF4A",
  "megakaryocytic_erythroid" = "#E41A1C", 
      "early_lymphoid_bcell" = "#377EB8",
                      "t_nk" = "#FF7F00"
)

types <- names(types_palette)
  
fractions <- "../data/cibersortx_results/CIBERSORTx_Job2_Results.csv" |> 
  read_csv(show_col_types = FALSE) |> 
  select(sample = Mixture, all_of(types)) 
```

# Exploratory plots

Barplots, with rows ordered by hierarchical clustering. 

```{r }
fractions_matrix <- fractions |> 
  column_to_rownames("sample") |> 
  as.matrix()  

clust <- fractions_matrix |> 
  dist() |> 
  hclust() 

hclust_order <- clust$order |> 
  set_names(clust$labels)

base_barplot <- sample_metadata |> 
  filter(!is.na(sub_group)) |> 
  select(sample, sub_group, who) |> 
  inner_join(fractions, by = "sample") |> 
  mutate(
    hclust_order = hclust_order[sample], 
    sample = fct_reorder(sample, hclust_order)
    ) |> 
  pivot_longer(all_of(types), names_to = "cell_type", values_to = "fraction") |> 
  mutate(cell_type = factor(cell_type, levels = types)) |> 
  ggplot(aes(y = sample, x = fraction, fill = cell_type)) +
  geom_col(color = NA) +
  scale_x_reverse() +
  scale_fill_manual(values = types_palette) +
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    strip.text.y = element_text(angle = 0), 
    strip.clip = "off", 
    plot.title = element_text(hjust = 1),
    plot.title.position = "plot", 
    legend.position = "bottom",
    axis.title.y = element_blank()
    ) 
```


```{r fig.height=10, fig.width=10}
stacked_barplot_by_subgroup <- base_barplot +
  facet_grid(sub_group ~ . , scales = "free", space = "free") +
  ggtitle("Subgroup")

stacked_barplot_by_who <- base_barplot +
  facet_grid(who ~ . , scales = "free", space = "free") +
  ggtitle("WHO 2022 classification")

stacked_barplot <- plot_grid(
  # Plots
  plot_grid(
    stacked_barplot_by_subgroup + theme(legend.position = "none"),
    stacked_barplot_by_who  + theme(legend.position = "none"),
    ncol = 2
  ), 
  # Legend
  get_legend(stacked_barplot_by_subgroup), 
  ncol = 1, rel_heights = c(15, 1)
)
  
stacked_barplot
```


```{r fig.height=10, fig.width=10, message=FALSE, warning=FALSE}
ggsave(
  "../figures/stacked_barplot_cell_fractions.pdf",
  stacked_barplot,
  width = 10,
  height = 10,
  units = "in"
  )
```

# PCAs of the cell fractions

```{r fig.height=10, fig.width=10}

# I don't think fractions should be scaled

pca <- prcomp(fractions_matrix) 

rotation <- pca$rotation |> 
  as.data.frame() |> 
  rownames_to_column("cell_type") |> 
  mutate(cell_type = str_replace_all(cell_type, "_", "\n"))

importances <- summary(pca)$importance[2,] |> 
  signif(2)

base_pca <- pca$x |> 
  as.data.frame() |> 
  rownames_to_column("sample") |> 
  left_join(sample_metadata, by = "sample") |> 
  ggplot(aes(x = PC1, y = PC2)) +
  geom_segment(
    data = rotation, 
    aes(x = 0, y = 0, xend = PC1, yend = PC2)
  ) +
  geom_label(
    data = rotation, 
    size = 3,
    aes(x = PC1, y = PC2, label = cell_type)
  ) +
  xlab(glue("PC1 ({importances[1]})")) + 
  ylab(glue("PC2 ({importances[2]})")) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    )

```

## Colored by subgroup

```{r fig.height=10, fig.width=10}
pca_by_subgroup <- base_pca

pca_by_subgroup$layers <- c(
  geom_point(aes(color = sub_group)), 
  pca_by_subgroup$layers
)

pca_by_subgroup <- pca_by_subgroup +
  scale_color_manual(values = subgroup_palette)

pca_by_subgroup

ggsave(
  "../figures/pca_cell_fractions_by_subgroup.pdf",
  pca_by_subgroup,
  width = 10,
  height = 10,
  units = "in"
  )
```

* SF3B1 are enriched in erythroid. This is expected and acts as a sanity check. 

* del5qs are enriched in lymphocytes. These have an inflammatory biology so it checks out. 

## Highlight of newly defined groups

```{r fig.height=10, fig.width=10}
pca_by_subgroup_hl <- base_pca

pca_by_subgroup_hl$layers <- c(
  geom_point(
    data = pca_by_subgroup_hl$data |> 
      filter(sub_group %in% c("TET2-bi", "STAG2","-7", "EZH2")), 
    aes(color = sub_group),
    ), 
  pca_by_subgroup_hl$layers
)

pca_by_subgroup_hl <- pca_by_subgroup_hl +
  scale_color_manual(values = subgroup_palette) +
  scale_alpha_manual(values = 0:1)

pca_by_subgroup_hl

```

Focusing on the newly defined groups: 

* TET2bi is enriched in HPSCs and myeloid lineage, while depleted in erythroid/megakaryocytic. 

* STAG2 is enriched in HSPCs, and depleted in eveythig else "leukemic stem cell-like". 

## Colored by WHO 2022

```{r fig.height=10, fig.width=10}
pca_by_who <- base_pca
pca_by_who$layers <- c(geom_point(aes(color = who)), pca_by_who$layers)
pca_by_who <- pca_by_who + scale_fill_brewer(palette = "Set3")
pca_by_who

ggsave(
  "../figures/pca_cell_fractions_by_who.pdf",
  pca_by_who,
  width = 10,
  height = 10,
  units = "in"
  )
```

# Boxplots of cell fractions

```{r}
boxplots_df <- fractions |>
  inner_join(sample_metadata, by = "sample") |> 
  pivot_longer(all_of(types), values_to = "fraction", names_to = "cell_type") |> 
  pivot_longer(c(sub_group, who), names_to = "classification", values_to = "group") |> 
  mutate(
    cell_type = factor(cell_type, levels = types), 
    group = factor(group, levels = c(subgroups, "MDS-5q", "MDS-SF3B1", "MDS-TP53"))
    ) 
```

## Grouped by cell type


* TET2 bi 


* The newly defined STAG2 has a different profile from any of the previous groups, hspcs on top. 

```{r fig.height=7, fig.width=12}
boxplot_by_type <- boxplots_df |> 
  ggplot(aes(x = cell_type, color = cell_type, y = fraction)) +
  geom_boxplot(outlier.alpha = 0.5, outlier.size = 1) +
  facet_grid(classification ~ group) +
  scale_color_manual(values = types_palette) +
  theme(axis.text.x = element_blank(), legend.position = "bottom")

boxplot_by_type

ggsave(
  "../figures/boxplot_by_type_cell_fractions.pdf",
  boxplot_by_type,
  width = 12,
  height = 7,
  units = "in"
  )
```

```{r fig.height=7, fig.width=12}
boxplot_by_group <- fractions |>
  inner_join(sample_metadata, by = "sample") |> 
  pivot_longer(all_of(types), values_to = "fraction", names_to = "cell_type") |> 
  pivot_longer(c(sub_group, who), names_to = "classification", values_to = "group") |> 
  mutate(
    cell_type = factor(cell_type, levels = types), 
    group = factor(group, levels = c(subgroups, "MDS-5q", "MDS-SF3B1", "MDS-TP53"))
    ) |> 
  ggplot(aes(x = group, y = fraction, color = group)) +
  scale_color_manual(values = c(
    subgroup_palette,
    "MDS-5q" = "#F0E442", 
    "MDS-SF3B1" = "#0072B2",
    "MDS-TP53" = "#D55E00")
    ) +
  facet_grid(classification ~ cell_type, scales = "free", space = "free") +
  geom_boxplot(outlier.alpha = 0.5, outlier.size = 0.5) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "none",
  )
  
boxplot_by_group

ggsave(
  "../figures/boxplot_by_group_cell_fractions.pdf",
  boxplot_by_group,
  width = 12,
  height = 7,
  units = "in"
  )
```

# Dirichlet regression

First we evaluate if there are zeroes

```{r}
sum(fractions_matrix == 0)
```

There are, so we add a pseudocount

```{r}
pseudocount <- unique(sort(as.double(fractions_matrix)))[2]
pseudoc_matrix <- fractions_matrix + pseudocount
pseudoc_matrix <- pseudoc_matrix / rowSums(pseudoc_matrix)
```

## New subgroups 

We fit the Dirichlet using this modeling:

* We filter out the previously existing molecular subgroups (`del5q-IB`, `SF3B1-IB`, `Complex`). 

* We collapse the morphological groups (`MDS-LB`, `MDS-IB1`, `MDS-IB2`) into a single baseline subgroup. 

* We leave the newly defined molecular subgroups (`EZH2`, `TET2-bi`, `-7` and `STAG2`). 

* We fit the model cell fractions ~ new_subgroup 

We set the new collapsed `morphological` as the baseline group. 

```{r}
dr_df <- as.data.frame(pseudoc_matrix) |> 
  rownames_to_column("sample") |> 
  inner_join(select(sample_metadata, sample, sub_group), by = "sample") |> 
  select(sub_group, all_of(types)) |> 
  filter(!sub_group %in% c("del5q-IB", "SF3B1-IB", "Complex")) |> 
  mutate(
    sub_group = if_else(
      sub_group %in% c("MDS-LB", "MDS-IB1", "MDS-IB2"),
      "morphological",
      sub_group
      ),
    sub_group = fct_drop(fct_relevel(sub_group, "morphological"))
    )

Y <- DR_data(select(dr_df, -sub_group))

fit <- DirichReg(Y ~ sub_group, data = dr_df)

s <- summary(fit)

s
```

Coefficients of each subgroup vs clinical measurements. Only significant (pvalue < 0.05)
estimates are shown. 

```{r message=FALSE, warning=FALSE}
s$coef.mat |> 
  as.data.frame() |> 
  split(unlist(map2(s$n.vars, s$varnames, ~rep(.y, .x)))) |> 
  bind_rows(.id = "cell_type") |> 
  rownames_to_column("subgroup") |> 
  mutate(
    cell_type = factor(cell_type, levels = types), 
    subgroup = str_match(subgroup, "sub_group(.+)\\.\\d$")[,2], 
    Estimate = if_else(`Pr(>|z|)` < 0.05, Estimate, NA)
    ) |> 
  filter(subgroup != "X") |> 
  ggplot(aes(x = subgroup, y = cell_type, fill = Estimate)) +
  geom_tile() +
  geom_text(aes(label = signif(Estimate, 2))) +
  scale_fill_distiller(palette = "Spectral", na.value = "lightgrey") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

To compute confidence intervals for the predictions, we run 100 bootstraps of 
the modelling

```{r}
collapsed_subgroups <- levels(dr_df$sub_group)

predict_bootstrap_dr <- function(i){
  # Stratified subsampling with replacement
  df_boot <- dr_df |> 
    group_by(sub_group) |> 
    slice_sample(prop = 1, replace = TRUE) |> 
    ungroup() |> 
    drop_na()
  
  # Fit
  y_boot  <- DR_data(select(df_boot, -sub_group))
  fit     <- DirichReg(y_boot ~ sub_group, data = df_boot)
  
  # Predict and reformat
  pred <- predict(fit, tibble(sub_group = collapsed_subgroups)) 
  colnames(pred) <- types
  pred <- as.data.frame(pred)
  pred$sub_group <- factor(collapsed_subgroups, levels = collapsed_subgroups)
  pred
}

n_bootstraps <- 100

bootstraps_results <- 1:n_bootstraps |> 
  map(predict_bootstrap_dr, .progress = "bootstrapping") |> 
  bind_rows() |> 
  pivot_longer(all_of(types), names_to = "cell_type", values_to = "fraction") |> 
  group_by(sub_group, cell_type) |> 
  summarise(
    median = median(fraction), 
    lower_ci = quantile(fraction, 0.025),
    upper_ci = quantile(fraction, 0.975), 
    .groups = "drop_last"
    )  |> 
  ungroup() |> 
  mutate(cell_type = factor(cell_type, levels = types))
```

```{r}
dodge <- 0.5

bootstraps_results |> 
  ggplot() +
  geom_point(
    aes(x = cell_type, y = median, color = sub_group), 
    position = position_dodge(width = dodge)
    ) + 
  geom_segment(
    aes(x = cell_type, y = lower_ci, yend = upper_ci, color = sub_group), 
    position = position_dodge(width = dodge)
  ) +
  scale_color_manual(values = c(subgroup_palette, "morphological" = "black")) +
  ylab("Predicted cellular fraction\n (with 95% CIs)\n across 100 bootstraps") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
dodge <- 0.5

bootstraps_results |> 
  ggplot() +
  geom_point(
    aes(x = sub_group, y = median, color = cell_type), 
    position = position_dodge(width = dodge)
    ) + 
  geom_segment(
    aes(x = sub_group, y = lower_ci, yend = upper_ci, color = cell_type), 
    position = position_dodge(width = dodge)
  ) +
  scale_color_manual(values = types_palette) +
  ylab("Predicted cellular fraction\n (with 95% CIs)\n across 100 bootstraps") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Subgroup 

We fit the model cell fractions ~ subgroup

We set `MDS-LB` as the baseline group

```{r}
dr_df <- as.data.frame(pseudoc_matrix) |> 
  rownames_to_column("sample") |> 
  inner_join(select(sample_metadata, sample, sub_group), by = "sample") |> 
  select(sub_group, all_of(types)) |> 
  mutate(sub_group = fct_relevel(sub_group, "MDS-LB"))

Y <- DR_data(select(dr_df, -sub_group))

fit <- DirichReg(Y ~ sub_group, data = dr_df)

s <- summary(fit)

s
```

Coefficients of each subgroup vs clinical measurements. Only significant (pvalue < 0.05)
estimates are shown. 

```{r message=FALSE}
s$coef.mat |> 
  as.data.frame() |> 
  split(unlist(map2(s$n.vars, s$varnames, ~rep(.y, .x)))) |> 
  bind_rows(.id = "cell_type") |> 
  rownames_to_column("subgroup") |> 
  mutate(
    cell_type = factor(cell_type, levels = types), 
    subgroup = str_match(subgroup, "sub_group(.+)\\.\\d$")[,2], 
    Estimate = if_else(`Pr(>|z|)` < 0.05, Estimate, NA)
    ) |> 
  filter(subgroup != "X") |> 
  ggplot(aes(x = subgroup, y = cell_type, fill = Estimate)) +
  geom_tile() +
  geom_text(aes(label = signif(Estimate, 2))) +
  scale_fill_distiller(palette = "Spectral", na.value = "lightgrey") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## WHO 2022

We use `MDS-LB` as the baseline group

```{r}
dr_df <- as.data.frame(pseudoc_matrix) |> 
  rownames_to_column("sample") |> 
  left_join(select(sample_metadata, sample, who), by = "sample") |> 
  select(who, all_of(types)) |> 
  mutate(who = fct_relevel(who, "MDS-LB"))

Y <- DR_data(select(dr_df, -who))

fit <- DirichReg(Y ~ who, data = dr_df)

s <- summary(fit)
```

```{r}
s$coef.mat |> 
  as.data.frame() |> 
  split(unlist(map2(s$n.vars, s$varnames, ~rep(.y, .x)))) |> 
  bind_rows(.id = "cell_type") |> 
  rownames_to_column("who") |> 
  mutate(
    cell_type = factor(cell_type, levels = types), 
    who = str_match(who, "who(.+)\\.\\d$")[,2], 
    Estimate = if_else(`Pr(>|z|)` < 0.05, Estimate, NA)
    ) |> 
  filter(who != "X") |> 
  ggplot(aes(x = who, y = cell_type, fill = Estimate)) +
  geom_tile() +
  geom_text(aes(label = signif(Estimate, 2))) +
  scale_fill_distiller(palette = "Spectral", na.value = "lightgrey") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Hematogram

We predict the hematological clinical variables, scaling them first with z-score

```{r}
hemato_vars <- c("bm_blast","wbc","hb","plt")

hematogram <- sample_metadata |> 
  select(sample, all_of(hemato_vars)) |> 
  drop_na() |> 
  column_to_rownames("sample") |> 
  as.matrix() |> 
  scale() |> 
  as.data.frame() |> 
  rownames_to_column("sample")

dr_df <- as.data.frame(pseudoc_matrix) |> 
  rownames_to_column("sample") |> 
  left_join(hematogram, by = "sample")

Y <- DR_data(select(dr_df, -all_of(hemato_vars), -sample))

fit <- DirichReg(
  as.formula(glue("Y ~ {paste(hemato_vars, collapse = ' + ')}")),
  data = dr_df
  )

s <- summary(fit)

s
```

```{r warning=FALSE}
s$coef.mat |> 
  as.data.frame() |> 
  split(unlist(map2(s$n.vars, s$varnames, ~rep(.y, .x)))) |> 
  bind_rows(.id = "cell_type") |> 
  rownames_to_column("measure") |> 
  mutate(
    cell_type = factor(cell_type, levels = types), 
    measure = str_split_i(measure, "\\.", 1), 
    Estimate = if_else(`Pr(>|z|)` < 0.05, Estimate, NA)
    ) |> 
  filter(measure != "X") |> 
  ggplot(aes(x = measure, y = cell_type, fill = Estimate)) +
  geom_tile() +
  geom_text(aes(label = signif(Estimate, 2))) +
  scale_fill_distiller(palette = "Spectral", na.value = "lightgrey") 

```

Coefficients of each cell type vs clinical measurements. Only significant (pvalue < 0.05)
estimates are shown. 