---
title: "Analysis of cell samples"
date: "`r Sys.Date()`"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
pkgdown:
  as_is: true
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(glue)
library(readxl)
library(RColorBrewer)
library(cowplot)
library(SummarizedExperiment)
library(DirichletReg)
```

```{r}
set.seed(2026)
```


# Import data

```{r}
load("../data/raw/rnaseq_hersh.Rdata")

tet2_bi <- read_xlsx("../data/raw/20260305_TET2_MDS_share.xlsx") |> 
  filter(TET2bi == 1) |> 
  pull(ID)

who_levels <- c("MDS-TP53", "MDS-SF3B1",  "MDS-5q", "MDS-IB1",  "MDS-IB2",  "MDS-LB")
subgroups <- c( "Complex",  "SF3B1",      "del5q",  "MDS-IB1",  "MDS-IB2",  "Low blasts",
                "TET2 biallelic", "TET2 monoallelic", "STAG2","7-", "EZH2")

sample_metadata <- colData(se_hersh) |> 
  as.data.frame() |> 
  select(-SEX, -AGE, -Exam_Array) |> 
  rename_with(str_to_snake) |> 
  rownames_to_column("sample") |> 
  select(sample, who, bm_blast, wbc, hb, plt, ipssm, sub_group) |> 
  mutate(who = factor(who, levels = who_levels)) |> 
  filter(!is.na(who)) |> 
  mutate(
    sub_group = case_when(
      sub_group != "Unknown TET2" ~ sub_group, 
      sub_group ==  "Unknown TET2" & sample %in% tet2_bi ~ "TET2 biallelic", 
      sub_group == "Unknown TET2" & !sample %in% tet2_bi ~ "TET2 monoallelic"
    ), 
    sub_group = factor(sub_group, levels = subgroups)
  )
```





```{r}
types_palette <- c(
  "hspc" = "#984EA3", 
  "dc_monocytic_gmp" = "#4DAF4A" ,
  "megakaryocytic_erythroid" = "#E41A1C", 
  "early_lymphoid_bcell" = "#377EB8",
  "t_nk" = "#FF7F00"
)

types <- names(types_palette)
  
fractions <- Sys.glob("../data/cibersortx_results/CIBERSORTx_Job2_Results.csv") |> 
  read_csv(show_col_types = FALSE) |> 
  select(sample = Mixture, all_of(types)) 

table(as.logical(as.matrix(column_to_rownames(fractions, "sample")) == 0))
```

# Exploratory plots

Barplots, with rows ordered by hierarchical clustering. 

```{r }
fractions_matrix <- fractions |> 
  column_to_rownames("sample") |> 
  as.matrix()  

clust <- fractions_matrix |> 
  dist() |> 
  hclust() 

hclust_order <- clust$order |> 
  set_names(clust$labels)

base_barplot <- sample_metadata |> 
  filter(!is.na(sub_group)) |> 
  select(sample, sub_group, who) |> 
  inner_join(fractions, by = "sample") |> 
  mutate(
    hclust_order = hclust_order[sample], 
    sample = fct_reorder(sample, hclust_order)
    ) |> 
  pivot_longer(all_of(types), names_to = "cell_type", values_to = "fraction") |> 
  mutate(cell_type = factor(cell_type, levels = types)) |> 
  ggplot(aes(y = sample, x = fraction, fill = cell_type)) +
  geom_col(color = NA) +
  scale_x_reverse() +
  scale_fill_manual(values = types_palette) +
  theme(
    axis.text.y = element_blank(), 
    axis.ticks.y = element_blank(), 
    strip.text.y = element_text(angle = 0), 
    strip.clip = "off", 
    plot.title = element_text(hjust = 1),
    plot.title.position = "plot", 
    legend.position = "bottom",
    axis.title.y = element_blank()
    ) 
```


```{r fig.height=10, fig.width=10}
stacked_barplot_by_subgroup <- base_barplot +
  facet_grid(sub_group ~ . , scales = "free", space = "free") +
  ggtitle("Subgroup")

stacked_barplot_by_who <- base_barplot +
  facet_grid(who ~ . , scales = "free", space = "free") +
  ggtitle("WHO 2022 classification")

stacked_barplot <- plot_grid(
  # Plots
  plot_grid(
    stacked_barplot_by_subgroup + theme(legend.position = "none"),
    stacked_barplot_by_who  + theme(legend.position = "none"),
    ncol = 2
  ), 
  # Legend
  get_legend(stacked_barplot_by_subgroup), 
  ncol = 1, rel_heights = c(15, 1)
)
  
stacked_barplot

ggsave(
  "../figures/stacked_barplot_cell_fractions.pdf",
  stacked_barplot,
  width = 10,
  height = 10,
  units = "in"
  )
```

# PCAs of the cell fractions

```{r fig.height=10, fig.width=10}
pca <- prcomp(fractions_matrix)

rotation <- pca$rotation |> 
  as.data.frame() |> 
  rownames_to_column("cell_type") |> 
  mutate(cell_type = str_replace_all(cell_type, "_", "\n"))

importances <- summary(pca)$importance[2,] |> 
  signif(2)

base_pca <- pca$x |> 
  as.data.frame() |> 
  rownames_to_column("sample") |> 
  left_join(sample_metadata, by = "sample") |> 
  ggplot(aes(x = PC1, y = PC2)) +
  geom_segment(
    data = rotation, 
    aes(x = 0, y = 0, xend = PC1, yend = PC2)
  ) +
  geom_label(
    data = rotation, 
    size = 3,
    aes(x = PC1, y = PC2, label = cell_type)
  ) +
  xlab(glue("PC1 ({importances[1]})")) + 
  ylab(glue("PC2 ({importances[2]})")) +
  theme_minimal() +
  theme(
    axis.text = element_blank(),
    )

```

Colored by subgroup

```{r fig.height=10, fig.width=10}
pca_by_subgroup <- base_pca

pca_by_subgroup$layers <- c(
  geom_point(aes(color = sub_group)), 
  pca_by_subgroup$layers
)

pca_by_subgroup <- pca_by_subgroup +
  scale_color_brewer(palette = "Set3")

pca_by_subgroup

ggsave(
  "../figures/pca_cell_fractions_by_subgroup.png",
  pca_by_subgroup,
  width = 10,
  height = 10,
  units = "in"
  )
```

* SF3B1 are enriched in erythroid. This is expected and acts as a sanity check. 

* del5qs are enriched in lymphocytes. These have an inflammatory biology so it checks out. 



```{r fig.height=10, fig.width=10}
pca_by_subgroup_hl <- base_pca

pca_by_subgroup_hl$layers <- c(
  geom_point(
    aes(
      color = sub_group, 
      alpha = sub_group %in% c("TET2 biallelic", "TET2 monoallelic", "STAG2","7-", "EZH2"), 
      ),
    ), 
  pca_by_subgroup_hl$layers
)

pca_by_subgroup_hl <- pca_by_subgroup_hl +
  scale_color_brewer(palette = "Set3") +
  scale_alpha_manual(values = 0:1) +
  theme(legend.position = "none")

pca_by_subgroup_hl

```


Focusing on the newly defined groups: 

* TET2bi is enriched in HPSCs and myeloid lineage, while depleted in erythroid/megakaryocytic. 

* STAG2 is enriched in HSPCs, and depleted in eveythig else "leukemic stem cell-like". 

Colored by WHO 2022

```{r fig.height=10, fig.width=10}
pca_by_who <- base_pca

pca_by_who$layers <- c(
  geom_point(aes(fill = who), color = "lightgrey", shape = 21), 
  pca_by_subgroup$layers
)

pca_by_who <- pca_by_who +
  scale_fill_brewer(palette = "Set3")

pca_by_who

ggsave(
  "../figures/pca_cell_fractions_by_who.png",
  pca_by_who,
  width = 10,
  height = 10,
  units = "in"
  )
```

# Boxplots of cell fractions

```{r fig.height=9, fig.width=9}

base_boxplot <- fractions |>
  inner_join(sample_metadata, by = "sample") |> 
  pivot_longer(all_of(types), values_to = "fraction", names_to = "cell_type") |> 
  mutate(cell_type = factor(cell_type, levels = types)) |> 
  ggplot(aes(x = cell_type, y = fraction, color = cell_type)) +
  geom_boxplot(outlier.alpha = 0.5, outlier.size = 0.5) +
  scale_color_manual(values = types_palette) +
  theme(
    axis.text.x = element_blank(), 
    strip.text.x = element_text(angle = 90, hjust = 1), 
    legend.position = "bottom",
  )

boxplot <- plot_grid(
  # Plots
  plot_grid(
    base_boxplot + facet_grid(~sub_group) +ggtitle("Subgroup") + theme(legend.position = "none"),
    base_boxplot + facet_grid(~who) + ggtitle("WHO") + theme(legend.position = "none"),
    ncol = 1
  ), 
  # Legend
  get_legend(base_boxplot), 
  ncol = 1, rel_heights = c(15, 1)
)
  
boxplot

ggsave(
  "../figures/boxplot_cell_fractions.pdf",
  boxplot,
  width = 9,
  height = 9,
  units = "in"
  )
```

```{r fig.height=7, fig.width=12}
boxplot_by_group <- fractions |>
  inner_join(sample_metadata, by = "sample") |> 
  pivot_longer(all_of(types), values_to = "fraction", names_to = "cell_type") |> 
  pivot_longer(c(sub_group, who), names_to = "classification", values_to = "group") |> 
  mutate(cell_type = factor(cell_type, levels = types)) |> 
  ggplot(aes(x = group, y = fraction, color = group)) +
  scale_color_manual(values = rep(brewer.pal(n = 9, "Set1"), 2)) +
  facet_grid(classification ~ cell_type, scales = "free", space = "free") +
  geom_boxplot(outlier.alpha = 0.5, outlier.size = 0.5) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    legend.position = "none",
  )
  
boxplot_by_group

ggsave(
  "../figures/boxplot_by_group_cell_fractions.pdf",
  boxplot_by_group,
  width = 12,
  height = 7,
  units = "in"
  )
```


# Dirichlet regression

First we evaluate if there are zeroes or NAs

```{r}
sum(fractions_matrix == 0)
```

```{r}
sum(is.na(fractions_matrix))
```

We add a pseudocount

```{r}
pseudocount <- unique(sort(as.double(fractions_matrix)))[2]
pseudoc_matrix <- fractions_matrix + pseudocount
pseudoc_matrix <- pseudoc_matrix / rowSums(pseudoc_matrix)
```

## Subgroup 

We set `Low blasts` as the baseline group. 

```{r}
dr_df <- as.data.frame(pseudoc_matrix) |> 
  rownames_to_column("sample") |> 
  inner_join(select(sample_metadata, sample, sub_group), by = "sample") |> 
  select(sub_group, all_of(types)) |> 
  mutate(sub_group = fct_relevel(sub_group, "Low blasts"))

Y <- DR_data(select(dr_df, -sub_group))

fit <- DirichReg(
  Y ~ sub_group,
  data = dr_df,
  control = list(iterlim = 1e9, tol1 = 1e-5, tol2 = 1e-10 )
)

summary(fit)
```

Bootstrap, to obtain confidence intervals. 

## WHO 2022

We use `MDS-LB` as the baseline group

```{r}
dr_df <- as.data.frame(pseudoc_matrix) |> 
  rownames_to_column("sample") |> 
  left_join(select(sample_metadata, sample, who), by = "sample") |> 
  select(who, all_of(types)) |> 
  mutate(who = fct_relevel(who, "MDS-LB"))

Y <- DR_data(select(dr_df, -who))

fit <- DirichReg(
  Y ~ who,
  data = dr_df, 
  control = list(iterlim = 1e9, tol1 = 1e-5, tol2 = 1e-10 )
)

summary(fit)
```

## Hematogram

We predict the hematological clinical variables, scaling them first with z-score

```{r}
hemato_vars <- c("bm_blast","wbc","hb","plt")

hematogram <- sample_metadata |> 
  select(sample, all_of(hemato_vars)) |> 
  drop_na() |> 
  column_to_rownames("sample") |> 
  as.matrix() |> 
  scale() |> 
  as.data.frame() |> 
  rownames_to_column("sample")

dr_df <- as.data.frame(pseudoc_matrix) |> 
  rownames_to_column("sample") |> 
  left_join(hematogram, by = "sample")

Y <- DR_data(select(dr_df, -all_of(hemato_vars), -sample))

fit <- DirichReg(
  as.formula(glue("Y ~ {paste(hemato_vars, collapse = ' + ')}")),
  data = dr_df, 
  control = list(iterlim = 1e9, tol1 = 1e-5, tol2 = 1e-10 )
)

s <- summary(fit)

s
```

```{r}
s$coef.mat |> 
  as.data.frame() |> 
  split(unlist(map2(s$n.vars, s$varnames, ~rep(.y, .x)))) |> 
  bind_rows(.id = "cell_type") |> 
  rownames_to_column("measure") |> 
  mutate(
    cell_type = factor(cell_type, levels = types), 
    measure = str_split_i(measure, "\\.", 1), 
    Estimate = if_else(`Pr(>|z|)` < 0.05, Estimate, NA)
    ) |> 
  filter(measure != "X") |> 
  ggplot(aes(x = measure, y = cell_type, fill = Estimate)) +
  geom_tile() +
  geom_text(aes(label = signif(Estimate, 2))) +
  scale_fill_distiller(palette = "Spectral") 

```

Coefficients of each cell type vs clinical measurements. Only significant (pvalue < 0.05)
estimates are shown. 